---
title: "Assignment_2"
author: "Mohtasim Hadi Rafi"
date: "2024-05-23"
output:
  html_document: default
  word_document: default
---

**Directions:** Use the template [**assignment3.Rmd**](https://auburn.instructure.com/courses/1574122/files/242631409?wrap=1 "assignment3.Rmd") for Problem 1 and use a script for Problem 2. Submit the following three files: the HTML file generated by the R Notebook file for Problem 1, the R Script file you created for Problem 2, and the file "bottles_of_beer.txt" that you generated with that script.

**Problem 1.** Write a function that will calculate the value of the expression $$
\frac{\hat{𝑝_{1}}−\hat{𝑝_{2}}}{\sqrt{\hat{p}(1-\hat{p})(\frac{1}{n_1}+\frac{1}{n_2})}} \text{ where } \hat{p} = \frac{x_{1}+x_2}{n_{1}+n_2}, \hat{p_1}=\frac{x_1}{n_1}, \hat{p_2}=\frac{x_2}{n_2}
$$ for given values of $x_1, x_2, n_1, \text{ and } n_2.$

What value do you get for $x_1=12, x_2=15, n_1=30, \text{ and } n_2=40?$ Square your result and compare with the value from the following R code:

```{r error = TRUE}
prop.test(x=c(12, 15), n=c(30, 40), correct=FALSE)$statistic
```

**Problem 2.** Write an R script that will write the complete lyrics of the song "99 bottles of beer on the wall" into a text file called "bottles_of_beer.txt".

Here are some hints. First change your working directory to the desktop so it will be easy to find the file that you are going to create. Did you do it right? You can check on the location of the working directory as follows.

```{r error=TRUE}
getwd() # check on location of the working directory
```

Next you will need to tell R that all output should be directed to the file "bottles_of_beer.txt".

```{r error=TRUE}
sink(file="bottles_of_beer.txt") # directs all output to the file "bottles_of_beer.txt"
```

If you don't know the lyrics of this song then look it up on Wikipedia. Next, you will need to write a loop. Here is a loop to get you started.

```{r error=TRUE}
for (i in 10:2){
  beer.on.the.wall <- paste(i, "bottles of beer on the wall\n")
  cat(beer.on.the.wall)
}
```

Finally, you need to tell R that output should be written again to the R console.

```{r error=TRUE}
sink() # from now on all output will be written again to the R console
```

Open the file "bottles_of_beer.txt" to see the result. This should get you started.

Instead of using the functions sink() and cat() you could also use the functions file(), writeLines(), and close(). The function file() is used to open the file for writing, writeLines() is used for writing lines to the file, and close() is used to close the file connection.
